Inferring the effective start dates of non-pharmaceutical interventions during \covid outbreaks

### 2021-01-14 ms/ms.tex
started revising for BMC

\author[1]{Ilia Kohanovski}
\author[2,3]{Uri Obolski}
\author[1,a]{Yoav Ram}

\affil[1]{School of Computer Science, Interdisciplinary Center Herzliya, Herzliya 4610101, Israel}
\affil[2]{School of Public Health, Tel Aviv University, Tel Aviv 6997801, Israel}{
\affil[3]{Porter School of the Environment and Earth Sciences, Tel Aviv University, Tel Aviv 6997801, Israel}
\affil[a]{Corresponding author: yoav@yoavram.com, ORCID 0000-0002-9653-4458}

@@

During Feb-Apr 2020, many countries implemented non-pharmaceutical interventions, such as school closures and lockdowns, with variable schedules, to control the \covid pandemic caused by the \sars virus.
Overall, these interventions seem to have successfully reduced the spread of the pandemic.
We hypothesise that the official and effective start date of such interventions can significantly differ, for example due to slow adoption by the population, or because the authorities and the public are unprepared.
We fit an SEIR model to case data from~12 countries to infer the effective start dates of interventions and contrast them with the official dates.
We find mostly late, but also early effects of interventions. For example, Italy implemented a nationwide lockdown on Mar~11, but we infer the effective date on Mar~17 ($\pm$2.99 days 95\% CI). In contrast, Germany announced a lockdown on Mar~22, but we infer an effective start date on Mar~19 ($\pm$ 1.05 days 95\% CI).
We demonstrate that differences between the official and effective start of NPIs can distort conclusions about their impact, and discuss potential causes and consequences of our results.
\end{abstract}

@@

In the following, we describe our findings in more detail.

@@

%%% Exact
\paragraph*{Exact effective start of NPIs.}
We find one case in which the official and effective dates match "\emph{like a Swiss watch}" and the credible interval is narrow.
Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
Indeed, the posterior median $\hat{\tau}$ is Mar~19 ($\pm$2.51 days 95\% CI, see \autoref{fig:early}). It's also worth mentioning that Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.

@@

In most investigated countries we find a late effective start of NPIs.
For example, in Italy and in Wuhan, China, the effective start of the lockdowns seems to have occurred five or more days after the official date (\autoref{fig:late}).
This difference might be explained, in some cases, by low compliance: In Italy, for example, the government plan to implement a lockdown in the Northern provinces leaked to the public, resulting in people leaving these provinces before the lockdown started~\citep{Gatto2020}.
Late effect of NPIs may also be due to the time required by both the government and the citizens to prepare for a lockdown, and for new guidelines to be adopted by the population.

@@

The likelihood function is defined in \autoref{eq:likelihood}.
The posterior distribution of the model parameters $P(\theta \mid \vec{X})$ is estimated using the affine-invariant ensemble sampler for Markov chain Monte Carlo (MCMC)~\citep{Goodman2010} implemented in the \texttt{emcee} Python package~\citep{Foreman-Mackey2013}.

We defined the following prior distributions on the model parameters $P(\theta)$: 
\begin{equation} \label{eq:priors}
\begin{aligned} 
Z & \sim \mathit{Uniform}(2, 5) \\
D & \sim \mathit{Uniform}(2, 5) \\
\mu & \sim \mathit{Uniform}(0.2, 1) \\
\beta & \sim \mathit{Uniform}(0.8, 1.5) \\
\lambda & \sim \mathit{Uniform}(0, 1) \\
\alpha_1, \alpha_2 & \sim \mathit{Uniform}(0.02, 1)\\
E(0) & \sim \mathit{Uniform}(0, 3000) \\
I_u(0) & \sim \mathit{Uniform}(0, 3000) \\
\Delta t & \sim \mathit{Uniform}(1, 5) \\
\tau &\sim \mathit{TruncatedNormal}\Big(\frac{\tau^*+\tau^0}{2}, \frac{\tau^*-\tau^0}{2}, 5, T-2\Big),
\end{aligned}
\end{equation}
where the prior for $\tau$ is a truncated normal distribution shaped so that the date of the first and last NPI, $\tau^0$ and $\tau^*$ (\autoref{table:NPI_dates}), are at minus and plus one standard deviation, and taking values only between 5 and  $T-2$, where $T$ is the number of days in the data $\vec{X}$.
We also tested an uninformative uniform prior, $\mathit{Uniform}(1,T-2)$.
WAIC (\autoref{eq:WAIC}) of a model with this uniform prior was either higher, or lower by less than 2, compared to WAIC of a model with the truncated normal prior.
The uninformative prior resulted in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom. This was probably due to MCMC chains being stuck in low posterior regions of the parameter space.
We therefore decided to use the more informative truncated normal prior for $\tau$.
Other priors follow \citet{Li2020}, with the following exceptions.
$\lambda$ is used to ensure transmission rates are lower after the start of the NPIs ($\lambda < 1$).
We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood and unreasonable parameter estimates, and therefore chose $\mathit{Uniform}(1,5)$ as the prior for $\Delta t$.
We also tried to estimate the value of $p_t$ before and after $\tau$, instead of keeping it fixed at $1/9$ and $1/6$. The model with fixed values was supported by the evidence (lower WAIC, see below) in 9 of 12 countries. Moreover, the estimates for Wuhan, China were $1/9$ and $1/6$, as in \citet{Li2020}. \\
Model fitting was calibrated for case data up to Mar~28, and then applied to data up to Apr~11.

@@

@@

\section*{References}
\nolinenumbers
\bibliographystyle{agsm}
%\bibliography{/Users/yoavram/Documents/library}
\bibliography{ms}

>>>>

\author[1]{Ilia Kohanovski}
\author[2,3]{Uri Obolski}
\author[1,4,a]{Yoav Ram}

\affil[1]{School of Computer Science, Interdisciplinary Center Herzliya, Herzliya 4610101, Israel}
\affil[2]{School of Public Health, Tel Aviv University, Tel Aviv 6997801, Israel}{
\affil[3]{Porter School of the Environment and Earth Sciences, Tel Aviv University, Tel Aviv 6997801, Israel}
\affil[4]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv 6997801, Israel}
\affil[a]{Corresponding author: yoav@yoavram.com, ORCID 0000-0002-9653-4458}

@@

During Feb-Apr 2020, many countries implemented non-pharmaceutical interventions, such as school closures and lockdowns, with variable schedules, to control the \covid pandemic caused by the \sars virus.
Overall, these interventions seem to have reduced the spread of the pandemic.
We hypothesise that the official and effective start date of such interventions can significantly differ, for example due to slow adoption by the population, or because the authorities and the public are unprepared.
We fit an SEIR model to case data from~12 regions to infer the effective start dates of interventions and contrast them with the official dates.
We find mostly late, but also early effects of interventions. For example, Italy implemented a nationwide lockdown on Mar~11, but we infer the effective date on Mar~17 ($\pm$2.99 days 95\% CI). In contrast, Germany announced a lockdown on Mar~22, but we infer an effective start date on Mar~19 ($\pm$ 1.05 days 95\% CI).
We demonstrate that differences between the official and effective start of NPIs can distort conclusions about their impact, and discuss potential causes and consequences of our results.
\end{abstract}

@@

@@

%%% Exact
\paragraph*{Exact effective start of NPIs.}
We find one case in which the official and effective dates match and the credible interval is narrow.
Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
Indeed, the posterior median $\hat{\tau}$ is Mar~19 ($\pm$2.51 days 95\% CI, see \autoref{fig:early}). It's also worth mentioning that Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.

@@

In most investigated regions we find a late effective start of NPIs.
For example, in Italy and in Wuhan, China, the effective start of the lockdowns seems to have occurred five or more days after the official date (\autoref{fig:late}).
This difference might be explained, in some cases, by low compliance: In Italy, for example, the government plan to implement a lockdown in the Northern provinces leaked to the public, resulting in people leaving these provinces before the lockdown started~\citep{Gatto2020}.
Late effect of NPIs may also be due to the time required by both the government and the citizens to prepare for a lockdown, and for new guidelines to be adopted by the population.

@@

The likelihood function is defined in \autoref{eq:likelihood}.
The posterior distribution of the model parameters $P(\theta \mid \vec{X})$ is estimated using the affine-invariant ensemble sampler for Markov chain Monte Carlo (MCMC)~\citep{Goodman2010} implemented in the \texttt{emcee} Python package~\citep{Foreman-Mackey2013}. Reported credible intervals, or CI, are calculated as the highest density intervals, or HDI~\citep{Kruschke2015}.

We defined the following prior distributions on the model parameters $P(\theta)$: 
\begin{equation} \label{eq:priors}
\begin{aligned} 
Z & \sim \mathit{Uniform}(2, 5) \\
D & \sim \mathit{Uniform}(2, 5) \\
\mu & \sim \mathit{Uniform}(0.2, 1) \\
\beta & \sim \mathit{Uniform}(0.8, 1.5) \\
\lambda & \sim \mathit{Uniform}(0, 1) \\
\alpha_1, \alpha_2 & \sim \mathit{Uniform}(0.02, 1)\\
E(0) & \sim \mathit{Uniform}(0, 3000) \\
I_u(0) & \sim \mathit{Uniform}(0, 3000) \\
\Delta t & \sim \mathit{Uniform}(1, 5) \\
\tau &\sim \mathit{TruncatedNormal}\Big(\frac{\tau^*+\tau^0}{2}, \frac{\tau^*-\tau^0}{2}, 5, T-2\Big),
\end{aligned}
\end{equation}
where the prior for $\tau$ is a truncated normal distribution shaped so that the date of the first and last NPI, $\tau^0$ and $\tau^*$ (\autoref{table:NPI_dates}), are at minus and plus one standard deviation, and taking values only between 5 and  $T-2$, where $T$ is the number of days in the data $\vec{X}$.
We also tested an uninformative uniform prior, $\mathit{Uniform}(1,T-2)$.
WAIC (\autoref{eq:WAIC}) of a model with this uniform prior was either higher, or lower by less than 2, compared to WAIC of a model with the truncated normal prior.
The uninformative prior resulted in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom. 
%This was probably due to MCMC chains being stuck in low posterior regions of the parameter space.
We therefore decided to use the more informative truncated normal prior for $\tau$.
Other priors follow \citet{Li2020}, with the following exceptions.
$\lambda$ is used to ensure transmission rates are lower after the start of the NPIs ($\lambda < 1$).
We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood and unreasonable parameter estimates, and therefore chose $\mathit{Uniform}(1,5)$ as the prior for $\Delta t$.
We also tried to estimate the value of $p_t$ before and after $\tau$, instead of keeping it fixed at $1/9$ and $1/6$. The model with fixed values was supported by the evidence (lower WAIC, see below) in 9 of 12 countries. Moreover, the estimates for Wuhan, China were $1/9$ and $1/6$, as in \citet{Li2020}. \\

@@

Model fitting was calibrated for case data up to Mar~28, and then applied to data up to Apr~11.

@@

\section*{References}
\nolinenumbers
\bibliographystyle{agsm}
\bibliography{/Users/yoavram/Documents/library}
%\bibliography{ms}
